Clustering/Orthofinder analysis

### 2022-03-08 analysis/Orthofinder.Rmd
split aquatics up

```{r}
per_group_spec <-
  per_spec %>% mutate(group = case_when(
    name %in% c(
      'A. trichopada',
      'B. distachyon',
      'O. sativa',
      'P. patens',
      'P. trichocarpa',
      'S. moellendorffii',
      'A. thaliana',
      'T. parvula',
      'V. vinifera',
      'Z. mays'
    ) ~ 'Terrestrials',
    name %in% c(
      'C. reinhardtii',
      'L. gibba',
      'S. polyrhiza',
      'O. lucimarinus',
      'W. australis'
    ) ~ 'Aquatics',
    name %in% c('P. australis', 'Z. muelleri', 'Z. marina', 'A. antarctica') ~ 'Seagrasses'
  )) 
```

@@

MUCH better. The biggest chunk of orthogroups is not in aquatics or seagrasses with 13562 orthogroups. 3437 orthogroups are unique to aquatics, but not in seagrasses - 1183 orthogroups are unique in seagrasses, but not in aquatics. 404 + 3437 + 2217 orthogroups are not in terrestrials.

OK now we need to do the GOenrichment for those three groups.

Let's also make an Venn plot, we have only three groups:
```{r fig.width=8}

@@

```{r}
gene_per_spec <- groups %>% pivot_longer(-Orthogroup) %>% 
  filter(!is.na(value)) %>% 
  mutate(group = case_when(
    name %in% c(
      'A. trichopada',
      'B. distachyon',
      'O. sativa',
      'P. patens',
      'P. trichocarpa',
      'S. moellendorffii',
      'A. thaliana',
      'T. parvula',
      'V. vinifera',
      'Z. mays'
    ) ~ 'Terrestrials',
    name %in% c(
      'C. reinhardtii',
      'L. gibba',
      'S. polyrhiza',
      'O. lucimarinus',
      'W. australis'
    ) ~ 'Aquatics',
    name %in% c('P. australis', 'Z. muelleri', 'Z. marina', 'A. antarctica') ~ 'Seagrasses'
  )) 
```

@@

```{r}
present_aquatics_only <- setdiff(setdiff(mylistgroup$Aquatics, mylistgroup$Terrestrials), 
                         mylistgroup$Seagrasses)
present_seagrasses_only <- setdiff(setdiff(mylistgroup$Seagrasses, mylistgroup$Aquatics), 
                           mylistgroup$Terrestrials)
present_terrestrials_only <- setdiff(setdiff(mylistgroup$Terrestrials, mylistgroup$Aquatics),
                             mylistgroup$Seagrasses)
print(cbind(length(present_aquatics_only), length(present_seagrasses_only), length(present_terrestrials_only)))

```
OK those numbers fit with the above Venn diagram. I now have three lists of orthogroup names.

@@

```{r}
present_seagrass_and_aquatics_only <- setdiff(intersect(mylistgroup$Aquatics, mylistgroup$Seagrasses), 
        mylistgroup$Terrestrials)
print(length(present_seagrass_and_aquatics_only))
```
That's the tiny intersection of genes shared between aquatics and seagrasses, let's also get the union of seagrass genes, aquatic genes, and their intersection

```{r}
present_seagrass_and_aquatics_union <- union(mylistgroup$Aquatics, mylistgroup$Seagrasses)
print(length(present_seagrass_and_aquatics_union))
```

@@

```{r}
present_all <- intersect(intersect(mylistgroup$Terrestrials, mylistgroup$Aquatics), mylistgroup$Seagrasses)
length(present_all)
```

@@

```{r}
# gene_per_spec has the genes per orthogroup, we need to join with the various set-results we've made
genes_in_all <- gene_per_spec %>% filter(Orthogroup %in% present_all) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_aquatics_and_seagrasses <- gene_per_spec %>% filter(Orthogroup %in% present_seagrass_and_aquatics_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_present_seagrass_and_aquatics_union <- gene_per_spec %>% filter(Orthogroup %in% present_seagrass_and_aquatics_union) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

genes_only_in_aquatics <- gene_per_spec %>% filter(Orthogroup %in% present_aquatics_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_seagrasses <- gene_per_spec %>% filter(Orthogroup %in% present_seagrasses_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_terrestrials <- gene_per_spec %>% filter(Orthogroup %in% present_terrestrials_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
```

@@

fileConn<-file('data/Lost_present_gene_lists/Genes_only_in_Aquatics.txt')
writeLines(genes_only_in_aquatics, fileConn)
close(fileConn)

@@

seagrass_venn <- ggplot() +
    geom_sf(aes_string(fill="count"), data = data@region) +
    geom_sf(aes_string(color = "id"), data = data@setEdge, show.legend = F,
            lty = 'solid', size = 1, color='gray') +
    geom_sf_text(aes_string(label = "name"), data = data@setLabel,
                 size = NA, fontface='italic',
                 color = 'black') +
    theme_void() +
     geom_sf_label(aes_string(label='both'),
                             data = region_label,
                             alpha= 0,
                             color = 'black',
                             size = NA,
                             lineheight = 0.85,
                             label.size = NA) +
  scale_fill_gradientn(colours=wes_palette("Zissou1", 100, type = "continuous")) + 
  theme(legend.position = "none") + 
  scale_x_continuous(expand = expansion(mult = .2)) #trick from https://github.com/gaospecial/ggVennDiagram/blob/9435aa0ab4abb470c670ecb938c71576461ccedc/vignettes/using-ggVennDiagram.Rmd#L65
seagrass_venn
```
# Both Venns together

```{r fig.height=10, fig.width=8}
group_venn / seagrass_venn +  plot_annotation(tag_levels = 'A')

>>>>

```{r}
per_group_spec <-
  per_spec %>% mutate(group = case_when(
    name %in% c(
      'A. trichopada',
      'B. distachyon',
      'O. sativa',
      'P. patens',
      'P. trichocarpa',
      'S. moellendorffii',
      'A. thaliana',
      'T. parvula',
      'V. vinifera',
      'Z. mays'
    ) ~ 'Terrestrials',
    name %in% c(
      'L. gibba',
      'S. polyrhiza',
      'W. australis'
    ) ~ 'Duckweeds',
    name %in% c(
      'C. reinhardtii',
      'O. lucimarinus'
    ) ~ 'Algae',
    name %in% c(
      'P. australis',
      'Z. muelleri',
      'Z. marina',
      'A. antarctica'
      ) ~ 'Seagrasses'
  )) 
```

@@

MUCH better. The biggest chunk of orthogroups is not in aquatics or seagrasses with 13562 orthogroups. 3845 orthogroups are in duckweeds, seagrasses, terrestrials, but not algae - 2217 orthogroups are unique in seagrasses.

OK now we need to do the GOenrichment for those four groups.

Let's also make an Venn plot, we have only four groups:
```{r fig.width=8}

@@

```{r}
gene_per_spec <- groups %>% pivot_longer(-Orthogroup) %>% 
  filter(!is.na(value)) %>% 
  mutate(group = case_when(
    name %in% c(
      'A. trichopada',
      'B. distachyon',
      'O. sativa',
      'P. patens',
      'P. trichocarpa',
      'S. moellendorffii',
      'A. thaliana',
      'T. parvula',
      'V. vinifera',
      'Z. mays'
    ) ~ 'Terrestrials',
    name %in% c(
      'L. gibba',
      'S. polyrhiza',
      'W. australis'
    ) ~ 'Duckweeds',
    name %in% c(
      'C. reinhardtii',
      'O. lucimarinus'
    ) ~ 'Algae',
    name %in% c(
      'P. australis',
      'Z. muelleri', 'Z. marina', 'A. antarctica') ~ 'Seagrasses'
  )) 
```

@@

```{r}
present_algae_only <- setdiff(setdiff(setdiff(mylistgroup$Algae, mylistgroup$Duckweeds), 
                         mylistgroup$Seagrasses), mylistgroup$Terrestrials)
present_seagrasses_only <- setdiff(setdiff(setdiff(mylistgroup$Seagrasses, mylistgroup$Duckweeds),                            mylistgroup$Terrestrials), mylistgroup$Algae)
present_terrestrials_only <- setdiff(setdiff(setdiff(mylistgroup$Terrestrials, mylistgroup$Duckweeds), mylistgroup$Seagrasses), mylistgroup$Algae)
present_duckweeds_only <- setdiff(setdiff(setdiff(mylistgroup$Duckweeds, mylistgroup$Terrestrials), mylistgroup$Seagrasses), mylistgroup$Algae)

print(cbind(length(present_algae_only), length(present_seagrasses_only), length(present_terrestrials_only), length(present_duckweeds_only)))

```
OK those numbers fit with the above Venn diagram. I now have four lists of orthogroup names.

@@

```{r}
present_seagrass_and_aquatics_only <- setdiff(intersect(intersect(mylistgroup$Duckweeds, mylistgroup$Seagrasses), mylistgroup$Algae),
        mylistgroup$Terrestrials)
print(length(present_seagrass_and_aquatics_only))
```
That's the tiny intersection of genes shared between aquatics and seagrasses, let's also get the union of seagrass genes, aquatic genes, and their intersection

```{r}
present_seagrass_and_aquatics_union <- union(mylistgroup$Algae, union(mylistgroup$Duckweeds, mylistgroup$Seagrasses))
print(length(present_seagrass_and_aquatics_union))
```

@@

```{r}
present_all <- intersect(intersect(mylistgroup$Algae, intersect(mylistgroup$Terrestrials, mylistgroup$Duckweeds)), mylistgroup$Seagrasses)
length(present_all)
```

@@

```{r}
# gene_per_spec has the genes per orthogroup, we need to join with the various set-results we've made
genes_in_all <- gene_per_spec %>% filter(Orthogroup %in% present_all) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_aquatics_and_seagrasses <- gene_per_spec %>% filter(Orthogroup %in% present_seagrass_and_aquatics_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

genes_present_seagrass_and_aquatics_union <- gene_per_spec %>% filter(Orthogroup %in% present_seagrass_and_aquatics_union) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

genes_only_in_duckweeds <- gene_per_spec %>% filter(Orthogroup %in% present_duckweeds_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_algae <- gene_per_spec %>% filter(Orthogroup %in% present_algae_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_seagrasses <- gene_per_spec %>% filter(Orthogroup %in% present_seagrasses_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_terrestrials <- gene_per_spec %>% filter(Orthogroup %in% present_terrestrials_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
```

@@

fileConn<-file('data/Lost_present_gene_lists/Genes_only_in_Duckweeds.txt')
writeLines(genes_only_in_duckweeds, fileConn)
close(fileConn)

fileConn<-file('data/Lost_present_gene_lists/Genes_only_in_Algae.txt')
writeLines(genes_only_in_algae, fileConn)
close(fileConn)

@@

seagrass_venn <- ggplot() +
    geom_sf(aes_string(fill="count"), data = data@region) +
    geom_sf(aes_string(color = "id"), data = data@setEdge, show.legend = F,
            lty = 'solid', size = 1, color='gray') +
    geom_sf_text(aes_string(label = "name"), data = data@setLabel,
                 size = NA, fontface='italic',
                 color = 'black') +
    theme_void() +
     geom_sf_label(aes_string(label='both'),
                             data = region_label,
                             alpha= 0,
                             color = 'black',
                             size = NA,
                             lineheight = 0.85,
                             label.size = NA) +
  scale_fill_gradientn(colours=wes_palette("Zissou1", 100, type = "continuous")) + 
  theme(legend.position = "none") + 
  scale_x_continuous(expand = expansion(mult = .2)) #trick from https://github.com/gaospecial/ggVennDiagram/blob/9435aa0ab4abb470c670ecb938c71576461ccedc/vignettes/using-ggVennDiagram.Rmd#L65
seagrass_venn
```

# Both Venns together

Let's save both Venns so I can combine them with other plots in other Rmds.

```{r}
save(group_venn, file='output/group_venn_image.Rdata')
save(seagrass_venn, file='output/seagrass_venn_image.Rdata')
```

```{r fig.height=10, fig.width=8}
group_venn / seagrass_venn +  plot_annotation(tag_levels = 'A')
```
